Bayesian Reconstruction of Coevolutionary Histories

### 2013-09-25 paper/article.tex
General progress

% Bibliography
\usepackage[american]{babel}
\usepackage[backend=biber,style=apa]{biblatex}
\DeclareLanguageMapping{american}{american-apa}
\bibliography{bibliography.bib}
\newcommand{\citeapos}[2]{\citeauthor{#1}'s #2 (\citeyear{#1})}
\usepackage{doi}

@@

Phylogenies, or evolutionary trees, can provide a valuable perspective to evolutionary processes. The topology of the tree depicts the set of speciation events that gave rise to those taxa and thus their ancestral relationships \parencite{Baum:2008}. Often, the lengths of branches indicate the number of substitutions (e.g., genetic mutations) accumulated or the amount of time passed at that particular lineage \parencite{Baum:2008}. By mapping traits such as physical attributes or geographic locations onto a known phylogeny, we can use the tree to learn about the mechanisms under which these traits change \parencites{Lemey:2009}{Lemey:2010}{Segraves:2010}.

Molecular sequences are the primary source of data for constructing phylogenies \parencite{Baum:2008}. With the advancements of DNA sequencing technology over the last few decades, considerable effort has been invested in developing robust statistical methods for inferring phylogenies from sequence alignments (of both nucleotides and amino acids) \parencite{Felsenstein:2005}. An early technique involved optimizing the tree topology under the parsimony criterion \parencite{Fitch:1971}. The most parsimonious, or optimal, tree is the \enquote{simplest} explanation for a sequence alignment---the tree involving the least number of substitutions. Following his criticism of parsimony methods \parencite{Felsenstein:1978}, \textcite{Felsenstein:1981} introduced an algorithm calculate the likelihood of a tree given a substitution model for evolution and sequence data. Similar to maximum parsimony, maximum likelihood techniques optimize a tree under Felsenstein's likelihood function, often simultaneously estimating parameter values for the substitution model \parencite{Felsenstein:2005}. The substitution model enables sophisticated models of evolution to be used to infer phylogenies, such as models aware of the underlying molecular evolutionary processes.

Many recent advances in phylogenetics have focused on methodologies utilizing Bayesian inference \parencites{Ronquist:2012}{Drummond:2012}. Bayesian inference takes its name from Bayes' theorem \parencite{Bayes:1763} and relies on the principle that the posterior probability of a model's parameters~$\theta$ given some data~$D$ is $P\left(\theta|D\right) \propto P\left(D|\theta\right) P\left(\theta\right)$ for a constant dataset. The likelihood, $P\left(D|\theta\right)$, is the probability of simulating the observed data under the model parameters (e.g., \citeauthor{Felsenstein:1981}'s \citeyear{Felsenstein:1981} algorithm), and the prior, $P\left(\theta\right)$, is the probability of the model parameters without considering the data. The prior can be a mechanism to apply previous knowledge to a new analysis; for example, constructing a prior distribution on the mutation rate representative of the rates reported in previous literature will \enquote{bias} the posterior to favor those rates. When explicit prior knowledge is not available, an uninformative prior may be applied.

Often we are interested in only some of the parameters involved in an analysis (e.g., just the tree topology), in which case we can calculate the posterior probability of specific values for those parameters by integrating over all possible values of the remaining \enquote{nuisance} parameters (branch lengths, mutation rate, etc.). Due to the complexity of the models, the resulting multi-dimensional integral cannot be evaluated directly but instead approximated via \ac{MCMC}. The Metropolis--Hastings algorithm \parencites{Metropolis:1953}{Hastings:1970} is a common implementation of the \ac{MCMC} method in phylogenetic inference

Although there have been criticisms of Bayesian \ac{MCMC} methodologies \parencites{Felsenstein:2005}{Kolaczkowski:2009}, they remain popular for \parencite{}.

@@

Given data~$D = \left(d_H,d_S,d_R\right)$, the host sequence data, the symbiont sequence data, and the leaf--leaf associations between the host and symbiont trees, respectively, I define the posterior probability of the host tree~$H$ and the symbiont tree~$S$, both with branch lengths, and the reconstruction~$R$ mapping internal nodes of $S$ to $H$ as
\begin{equation}
P\left(H,S,R|D\right) \propto \int P\left(d_S\right|S) P\left(S|H,R,d_R,\theta\right) P\left(H|d_H\right) P\left(R\right) P\left(\theta\right) \, \dif \theta
\label{eq:cophylogenyposterior}
\end{equation}
where parameters $\theta = \left(\lambda,\tau,\mu\right)$, the duplication, host-switch, and loss rates, respectively, and are integrated out. The probability of a tree~$T$ given sequence data~$D$ is $P\left(T|D\right) \propto P\left(D|T\right) P\left(T\right)$, where the likelihood is generally calculated with \citeapos{Felsenstein:1981} algorithm or a more complex multi-locus model, e.g. the multispecies coalescent \parencite{Heled:2010a}, and the prior with either a coalescent model \parencite{Kingman:1982} or a birth-death speciation model \parencite{Gernhard:2008}. In this case, the prior on the symbiont tree is represented instead by the reconstruction likelihood $P\left(S|H,R,d_R,\theta\right)$.

>>>>

% Bibliography
\usepackage[american]{babel}
\usepackage[backend=biber,style=apa]{biblatex}
\DeclareLanguageMapping{american}{american-apa}
\bibliography{bibliography.bib}
\newcommand{\aposcite}[2]{\citeauthor{#1}'s #2 (\citeyear{#1})}
\usepackage{doi}

@@

Phylogenies, or evolutionary trees, can provide a valuable perspective to evolutionary processes. The topology of a phylogenetic tree depicts the set of speciation events that gave rise to a set of taxa and thus their ancestral relationships \parencite{Baum:2008}. Often, the lengths of branches indicate the number of substitutions (e.g., genetic mutations) accumulated or the amount of time passed at that particular lineage \parencite{Baum:2008}. By mapping traits such as physical attributes or geographic locations onto a known phylogeny, we can use the tree to learn about the mechanisms under which these traits change \parencites{Lemey:2009}{Lemey:2010}{Segraves:2010}.

Molecular sequences are the primary source of data for constructing phylogenies \parencite{Baum:2008}. With the advancements of DNA sequencing technology over the last few decades, considerable effort has been invested in developing robust statistical methods for inferring phylogenies from sequence alignments (of both nucleotides and amino acids) \parencite{Felsenstein:2005}. One of the first techniques optimized the tree topology under the parsimony criterion \parencite{Fitch:1971}. The most parsimonious, or optimal, tree is the \enquote{simplest} possible explanation for a sequence alignment---the tree involving the least number of substitutions. Following his criticism of parsimony methods \parencite{Felsenstein:1978}, \textcite{Felsenstein:1981} introduced an algorithm to calculate the likelihood of a tree given a sequence data and a substitution model. Similar to maximum parsimony, maximum likelihood techniques optimize a tree under \aposcite{Felsenstein:1981}{likelihood function}, often simultaneously estimating parameter values for the substitution model \parencite{Felsenstein:2005}. Likelihood methods offer the advantage of sophisticated substitution models, such as models that consider the underlying molecular processes of evolution.

Many recent advances in phylogenetics have focused on methodologies utilizing Bayesian inference \parencites{Ronquist:2012}{Drummond:2012}. Bayesian inference takes its name from Bayes' theorem \parencite{Bayes:1763} and relies on the principle that the posterior probability of a model's parameters~$\theta$ given some constant data~$D$ is $P\left(\theta|D\right) \propto P\left(D|\theta\right) P\left(\theta\right)$. The likelihood, $P\left(D|\theta\right)$, is the probability of simulating the observed data under the model parameters (e.g., \citeauthor{Felsenstein:1981}'s \citeyear{Felsenstein:1981} algorithm), and the prior, $P\left(\theta\right)$, is the probability of the model parameters without considering the data. The prior can be a mechanism to apply previous knowledge to a new analysis; for example, constructing a prior distribution on the mutation rate representative of the rates reported in previous literature will \enquote{bias} the posterior to favor those rates. When explicit prior knowledge is not available, an uninformative prior may be applied.

Often we are interested in only some of the parameters involved in an analysis (e.g., just the tree topology), in which case we can calculate the posterior probability of specific values for those parameters by integrating over all possible values of the remaining \enquote{nuisance} parameters (branch lengths, mutation rate, etc.). Due to the complexity of the models, the resulting multi-dimensional integral cannot be evaluated directly but instead approximated via a \ac{MCMC}. The Metropolis--Hastings \ac{MCMC} algorithm advances the chain from one state to the next by randomly mutating the current state and accepting or rejecting the proposal based on its posterior probability. \parencites{Metropolis:1953}{Hastings:1970}. With a long enough chain, the collection of samples from the \ac{MCMC} is an approximation of the posterior distribution.

Although there have been criticisms of Bayesian \ac{MCMC} methodologies \parencites{Felsenstein:2005}{Kolaczkowski:2009} for phylogenetics, they remain popular for their flexibility  \parencites{Huelsenbeck:2000}{Ronquist:2012}.

@@

Given data~$D = \left(d_H,d_S,d_R\right)$, the host sequence data, the symbiont sequence data, and the leaf--leaf associations between the host and symbiont trees, respectively, I define the posterior probability of the host tree~$H$ and the symbiont tree~$S$, both with branch lengths, and the reconstruction~$R$ mapping internal nodes of $S$ to $H$ as
\begin{equation}
P\left(H,S,R|D\right) \propto \int P\left(d_S\right|S) P\left(S|H,R,d_R,\theta\right) P\left(H|d_H\right) P\left(R\right) P\left(\theta\right) \, \dif \theta
\label{eq:cophylogenyposterior}
\end{equation}
where parameters $\theta = \left(\lambda,\tau,\mu\right)$, the duplication, host-switch, and loss rates, respectively, and are integrated out. The probability of a tree~$T$ given sequence data~$D$ is $P\left(T|D\right) \propto P\left(D|T\right) P\left(T\right)$, where the likelihood is generally calculated with \aposcite{Felsenstein:1981} algorithm or a more complex multi-locus model, e.g. the multispecies coalescent \parencite{Heled:2010a}, and the prior with either a coalescent model \parencite{Kingman:1982} or a birth-death speciation model \parencite{Gernhard:2008}. In this case, the prior on the symbiont tree is represented instead by the reconstruction likelihood $P\left(S|H,R,d_R,\theta\right)$.
